Vowel Elision in Yoruba Collocations

### 2023-02-25 paper.tex
Add basic intro, some discussion on approach

@@

@@

We can describe these three distinct groups of data straightforwardly. In
Yoruba, when two vowels become adjacent via concatenation of Yoruba morphemes,
the first vowel is deleted, unless the second vowel is less sonorous,
in which case it is deleted. This deletion occurs in order to ensure that
syllables always have an onset. This generalization can be analyzed from an
optimality-theoretic viewpoint. To begin, the phenomena in
\pref{fig:deletion-simple} can be naturally accounted for with a prohibition
against onsetless syllables, so one of the conjoined vowels is deleted to
prevent such a syllable from surfacing. Specifically, a `V`, with any number of
`C`s after it, is not a valid candidate for a syllable, because it lacks an
onset. We can codify this observation via the constraint \onset, which
precisely moderates against candidate surface forms with syllables which lack
an onset. This is defined in \pref{def:onset-and-maxc}.

@@

The answer lies in Yoruba's preference for preserving morpheme-initial vowels
versus morpheme-final vowels. Cross-linguistically, onsets are more important
than codas-- looking to the corpus of Optimality Theory constraints, for
instance, we see that \onset{} exists to regulate against syllables
\textit{without} onsets, but \nocoda{} exists to regulate against syllables
\textit{with} codas. Given this cross-linguistic asymmetry, it is natural to
conclude that Yoruba's treatment of vowels at morpheme boundaries (which are
also syllable boundaries) follows a similar pattern.

@@

Our analysis so far accounts for the phenomena in \pref{fig:deletion-simple},
but not the phenomena in \pref{fig:deletion-except-1} or
\pref{fig:deletion-except-2}. We proceed, now, to analyze the first of these, in
which [i] is deleted regardless of whether it the morpheme initial or final
sound. One possible explanation as to why this is the case might be some
quality of the vowels themselves-- the frontness or backness of the vowels
appears to have no immediately discernible pattern with respect to deletion, so
we might try vowel height. Indeed, height appears to be correlated: [u] and
[i], which are both high vowels, are involved in the exceptional deletion
cases! We analyze both of these instances through the mechanism of sonority,
which offers an explanation for high vowels being involved in deletion-- they
are less sonorous. The sonority scale of vowels has been discussed in prior
work[\textbf{cite the quality-sensitive stress paper}, wikipedia article?].
This scale is shown in \pref{table:vowel-sonority}.

@@

Note that our winner-threatening loser could also have been created via a
featural change-- we can deal with this in a similar way, by introducing a
high-ranked \ident{} constraint. We show this below. For similar reasons to
that for \dep, \ident{} cannot be ranked against \onset. \dep{} and \ident{}
also cannot be ranked against each other, because we never see a surface
form that uses either.

@@

@@

>>>>

\usepackage{biblatex}
\addbibresource{references.bib}

@@

\section{Analysis}

@@

We can describe these three distinct groups of data straightforwardly. In
Yoruba, when two vowels become adjacent via concatenation of Yoruba morphemes,
the first vowel is deleted, unless the second vowel is less sonorous,
in which case it is deleted. This deletion occurs in order to ensure that
syllables always have an onset. This generalization can be analyzed from an
Optimality-theoretic viewpoint. To begin, the phenomena in
\pref{fig:deletion-simple} can be naturally accounted for with a prohibition
against onsetless syllables, so one of the conjoined vowels is deleted to
prevent such a syllable from surfacing. Specifically, a `V`, with any number of
`C`s after it, is not a valid candidate for a syllable, because it lacks an
onset. We can codify this observation via the constraint \onset, which
precisely moderates against candidate surface forms with syllables which lack
an onset. This is defined in \pref{def:onset-and-maxc}.

@@

The answer lies in Yoruba's preference for preserving morpheme-initial vowels
versus morpheme-final vowels. Cross-linguistically, onsets are more important
than codas--- looking to the corpus of Optimality Theory constraints, for
instance, we see that \onset{} exists to regulate against syllables
\textit{without} onsets, but \nocoda{} exists to regulate against syllables
\textit{with} codas. Given this cross-linguistic asymmetry, it is natural to
conclude that Yoruba's treatment of vowels at morpheme boundaries (which are
also syllable boundaries) follows a similar pattern.

@@

Our analysis so far accounts for the phenomena in \pref{fig:deletion-simple},
but not the phenomena in \pref{fig:deletion-except-1} or
\pref{fig:deletion-except-2}. We proceed, now, to analyze the first of these, in
which [i] is deleted regardless of whether it the morpheme initial or final
sound. One possible explanation as to why this is the case might be some
quality of the vowels themselves--- the frontness or backness of the vowels
appears to have no immediately discernible pattern with respect to deletion, so
we might try vowel height. Indeed, height appears to be correlated: [u] and
[i], which are both high vowels, are involved in the exceptional deletion
cases! We analyze both of these instances through the mechanism of sonority,
which offers an explanation for high vowels being involved in deletion--- they
are less sonorous. The sonority scale of vowels has been discussed in prior
work[\textbf{cite the quality-sensitive stress paper}, wikipedia article?].
This scale is shown in \pref{table:vowel-sonority}.

@@

Note that our winner-threatening loser could also have been created via a
featural change--- we can deal with this in a similar way, by introducing a
high-ranked \ident{} constraint. We show this below. For similar reasons to
that for \dep, \ident{} cannot be ranked against \onset. \dep{} and \ident{}
also cannot be ranked against each other, because we never see a surface
form that uses either.

@@

\pagebreak

@@

\pagebreak

\section{Computing Ideal Surface Forms}

\subsection{Introduction and Background}

We now turn to assigning a computational interpretation to our analysis.
Creating an algorithm to compute ideal outputs from Optimality-theoretic
constraints has been explored in prior work, though not extensively. Kartunnen
\cite{karttunen2006finite} explores a computational implementation of the
Optimality-theoretic analysis of Finnish stress using the finite state machine
description language XFST (\cite{karttunen2001applications} gives a broad
overview of finite state transducers as applied to morphological analysis and
parsing). Interestingly, such an implementation leads to insights about the
soundness of the preexisting analyses of Finnish stress--- namely, that they
are not. Key to this analysis is that it is an \textit{approximation}: indeed,
the paper itself notes that ``it has been known for a long time [\ldots] that
\textsc{OT} is not a finite-state system'', and that ``\textsc{OT} is
fundamentally more complex than [rule-based phonological systems]''.

Idsardi \cite{idsardi2006simple} explores this computational aspect, providing
a proof that generating an appropriate candidate set which can be evaluated
against some constraint ranking is \textit{NP-Hard}. This is a term used in
computability theory to describe a problem to which any number of other hard
problems can be \textit{reduced}. In essence, if there exists an efficient way
to solve this problem, a number of other incredibly difficult problems could
also be easily solved. Knowing a problem is NP-Hard is strong evidence that it
cannot be solved efficiently.

Heinz, Kobele, and Riggle \cite{heinz2009evaluating} contest this result,
showing that a different representation of \textsc{OT} grammars as a single
finite state transducer (instead of a sequence of finite state constriants)
averts the computational intractability issue demonstrated in
\cite{idsardi2006simple}. Additionally, they outline two subclasses of the
\textit{universal problem} of attempting to determine a valid constraint
ranking and underlying representations given only surface forms. Namely,
the \textit{simple problem} asks for optimal surface forms given an underlying
form and fixed constraint set \textsc{Con} and ranking \textsc{R}. The
\textsc{quasi-universal problem} asks for an optimal surface form given
a fixed constraint set with any possible ranking.

Encoding our analysis as an algorithm has several advantages, namely, that we
make explicit the procedure for calculating surface forms from underlying ones,
forcing us to make obvious any implicit assumptions made in the analysis as we
step through it. Secondly, and perhaps more interestingly, this enables us to
test our analysis quickly and thoroughly given new data, ensuring that it holds
and giving us opportunities to improve it otherwise.

\subsection{Approach}

We base our procedure, crucially, on bounded depth-first search. We first
describe the basic algorithm, before proceeding onwards to an implementation.
Viewed as a single function which takes an input and returns an output, our
analysis takes as input an underlying form and a set of ranked constraints,
and searches for the correct output form. The search for the correct output
is of course the crucial portion here, and is the part on which most of the
discussion will be spent.

Our basic approach employs \textit{bounded depth-first search}. It is best to
illustrate this technique on a concrete example. Take, for instance, the
surface form \textipa{[sol\'u]} and its underlying representation
\textipa{/se-ol\'u/}. We will demonstrate our technique on this mapping.

Our first order of business is to generate a reasonably large set of
permutations of \textipa{[sol\'u]}, one of which is the correct output form. We
are not yet attempting to identify which is the correct output form, only
generate it so it may be picked later. In effect, we are attempting to
efficiently emulate Optimality Theory's \textsc{Gen}. To do this, we intoduce
the following core transformations; we model each of these transformations as
functions of their own:

\begin{enumerate}
    \item \textbf{Delete}: Delete a segment
    \item \textbf{Epenthesize}: Insert a segment
    \item \textbf{Change}: Map a segment to another one
    \item \textbf{Metathesize}: Swap the positions of two segments
\end{enumerate}

\textbf{Delete} takes as input a form, such as \textipa{[sol\'u]}, and
randomly deletes one of the characters. For instance, it may delete
\textipa{[o]} to produce \textipa{[sl\'u]}, though again, this says little
about whether such a form would surface. It is more likely to delete inside the
word than at the edges of the word, following in line with broad phonological
generalizations about the markedness of deleting morpheme-initial or
morpheme-final segments (citation about this here?).

\textbf{Epenthesize} inserts a segment into its input and returns the new form.
For example, if given \textipa{[sol\'u]} as input, epenthesis may return
\textipa{[sool\'u]}. Again, it is more likely to epenthesize inside the form
than at its edges.

\textbf{Change}, to some extent, is functionally a composition of the previous
two operations, replacing one segment with another one. However, this does not
reflect how such transitions are modeled in the literature, using the
categorization of segments into their features. To accommodate this without
undertaking the modeling of autosegmental phonology
\cite{goldsmith1990autosegmental} in a computationally tractable way, we
manually partition the set of segments into phonologically similar (not
necessarily disjoint) sets, and attempt to map largely within these sets.
Illustrating concretely, we might classify all vowels in the language into one
set, and all the sonorants into another, so our two sets would look something
like this:

\subsection{Implementation}

\subsection{Analysis on corpus data}

\section{Concluding}

\printbibliography
